Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-12-28 ml.tex
for submission

    Collisions at the LHC can produce hundreds of
    particles with complex final-state configurations.
    This leads to the design of the ATLAS and CMS experiments
    to contain of the order 100 million detection elements
    in an attempt to disentangle these events. With these large
    arrays, terabytes of data are harvested from
    every collision. This presents a practical challenge for data
    collection where machine learning has could provide a solution.
    Indeed, boosted decision trees have been use
    to enhance triggers \cite{CMS:2020cmk} to accept or
    reject data before being saved to disk.
    % At the LHCb
    % experiment, 70\% of all data retained have had
    % some contact with machine learning algorithms \cite{LHCb:2014set}.

@@

    In a machine learning context, the fitting of a
    model to training data is termed training the model.
    Once the model is fitted, it is customary to use
    an unbiased dataset that is distinct from the training set, the test set,
    to evaluate the final model performance.

\subsection{Model of a neuron}
    In this context, a neuron is described by
    \begin{equation}\label{eqn:neuron}
        y = \phi(\mathbf{x}; \theta) = \phi(\mathbf{w}^{T}\mathbf{x} + b)
    \end{equation}
    where the model inputs, $\mathbf{x}$, are multiplied
    by the model weights, $\mathbf{w}$, before being summed
    with a bias term, $b$. Collectively, the model weights
    and biases are the model parameters, $\theta$.
    This combination of terms is then modified
    by an activation function, $\phi$, which is carefully
    chosen to perform a non-linear transformation. A neuron
    is illustrated in Figure~\ref{fig:neuron}.

@@

    The output of a densely-connected neural network can
    be formulated by repeatedly applying Eq. (\ref{eqn:neuron})
    to give
    \begin{equation}\label{eqn:nn_output}
        f(\mathbf{x}; \theta) = \boldsymbol{\phi}^{(n)}(\ldots \boldsymbol{\phi}^{(2)}(\; \boldsymbol{\phi}^{(1)}(\mathbf{x}; \theta^{(1)})\; ; \theta^{(2)}) \; ; \theta^{(n)}) \, ,
    \end{equation}
    where $\boldsymbol{\phi}^{(n)}$ and $\theta^{(n)}$
    denotes the outputs and parameters of layer $n$,
    where the output layer is included in this notation.
    By chaining together activation functions, and especially
    non-linear activation functions, neural networks
    become good function approximators
    \begin{equation}\label{eqn:nn_approx}
        f(\mathbf{x}; \theta) \approx g(\mathbf{x}) \, .
    \end{equation}
    The universal approximation theorem states that neural networks are
    able to represent any continuous function, $g(\mathbf{x})$,
    with one hidden layer and a finite number of neurons \cite{HORNIK1991251}.
    However, it is very difficult to achieve this due to
    practical constraints such as limited data and network size.
    It is much more common to link
    together a larger number of hidden layers, leading
    to deep neural networks since better learning algorithms
    have been found for this case \cite{lu2017expressive}.

@@

    In order to quantify the performance of a network for
    a given task it is useful to define a loss function.
    For regression tasks, such as approximating a function,
    a commonly used loss function is the mean squared error
    \begin{equation}\label{eqn:MSE}
        L_{\mathrm{MSE}} = \dfrac{1}{N} \sum_{i=0}^{N} (g(\mathbf{x}_{i}) - y(\mathbf{x}_{i}; \theta))^{2} \, ,
    \end{equation}
    where $g(\mathbf{x}_{i})$ and $f(\mathbf{x}_{i}; \theta)$
    are samples from the target distribution, and model predictions
    for inputs, $\mathbf{x}_{i}$, belonging to a data set with $N$
    samples. The loss function
    encodes the discrepancy between the truth value
    and the model prediction. Therefore, the task of finding
    an acceptable set of parameters can be reframed as an optimisation
    of the loss function. It should be noted that minimisation of
    the loss function is simply a proxy for maximising neural network
    predictive power. Due to the training and testing datasets being finite,
    the generalisation of the neural network is not guaranteed for the
    true underlying distribution from which the datasets are sampled
    from. This problem is generally referred to as overfitting.
    We will refer back to this problem in the context of
    fitting matrix elements in Section~\ref{sec:NN_ME}.

    Since neural networks are generally aimed at
    learning non-linear functions, the loss surface
    corresponding to Eq. (\ref{eqn:MSE}) is
    a function of many variables, and is likely highly
    non-convex with many local minima.
    The methods of choice for traversing these loss surfaces
    are all iterative gradient-based methods with modifications
    to improve convergence. Broadly speaking, the algorithms
    iteratively update the neural network parameters based on the
    gradient of the loss with respect to these parameters
    with some step size (or learning rate) $\eta$. That is
    \begin{equation}\label{eqn:gradient_descent}
        \theta \gets \theta - \eta \cdot \nabla_{\theta} L(\mathbf{x}_{\mathrm{batch}}; \theta) \, ,
    \end{equation}
    where $\nabla_{\theta}L(\mathbf{x}_{\mathrm{batch}}; \theta)$ is the
    gradient of the loss with respect to the parameters of the
    model, averaged over the batch of inputs $\mathbf{x}_{\mathrm{batch}}$.
    The initial state of the parameters
    are usually distributed according to normal
    or uniform distributions \cite{pmlr-v9-glorot10a,he_initialiser}.
    This variation of gradient descent is referred to
    as mini-batch gradient descent because the gradient
    $\nabla_{\theta}L$ is averaged over a mini-batch of
    samples $\mathbf{x}_{\mathrm{batch}}$. Mini-batch gradient
    descent strikes a balance between stochastic gradient
    descent (estimating gradient with one sample at a time),
    and batch gradient descent (calculating gradient with
    respect to entire training set) by having an efficient
    estimation of the gradient that is accurate enough
    for practical applications. In modern machine learning
    libraries, the updating of weights in Eq. (\ref{eqn:gradient_descent})
    is carried out via matrix multiplications which are
    highly efficient on graphics-processing units (GPUs),
    and the gradients are computed numerically with the aid
    of automatic differentiation tools. For these reasons, deep
    neural networks have become much more widespread due
    to the ease of constructing and training them, as well
    as their good predictive performance for general problems.

@@

    \subsection{Optimisation of hyperparameters}
    Another set of parameters that need to be optimised
    are the model hyperparameters.
    Hyperparameters are parameters that are
    not explicitly parameters of the model, instead they control
    the speed and quality of the training process.
    For instance, the learning rate in Eq. ({\ref{eqn:gradient_descent}}) is
    an important hyperparameter whose initial value has
    to be chosen carefully to observe a good rate of convergence.
    Some other examples would be
    the number of hidden layers, the choice of activation
    function, or the mini-batch size.

    The choice of hyperparameters is an important
    factor during the training process as a suboptimal set
    of hyperparameters can perform significantly worse
    than a more carefully chosen set. The process of tuning
    hyperparameters is often computationally expensive for two
    reasons. Firstly, the dimensionality of parameter space to choose from is large,
    and so the curse of dimensionality kicks in rapidly.
    Secondly, there is no exact a priori way to evaluate the
    final performance of a neural network given a set of
    hyperparameters, meaning training has to take place for
    some steps before the final model performance can be estimated.

@@

\section{Neural networks as matrix element surrogate models}\label{sec:NN_ME}
    Motivated by the ability of neural networks to
    model non-linear functions, as well as the rise
    in GPU usage in high energy physics, it seems
    natural to leverage neural networks to build surrogate
    models for matrix elements. There are multiple advantages
    to this: the time taken to evaluate matrix elements remains
    a bottleneck in event generation, and so building a
    fast surrogate model would accelerate this process.
    Once a fast and accurate surrogate model has been
    built, it can supplement traditional matrix
    element generators to increase statistical power of the sample
    by predicting on many more phase-space points. This
    requires the accumulated error in the surrogate model
    predictions to be lower than the statistical error in the
    integration of matrix elements, which is a key motivation
    for building an accurate surrogate model.

    The task of building a surrogate
    model for matrix elements is in principle straightforward.
    The dataset usually consists of phase-space kinematics
    in the form of four-momenta and the targets are the
    corresponding matrix elements for the process of interest,
    at the desired order in perturbation theory. As already
    discussed in Sections~\ref{sec:MC_integration} and \ref{sec:me_generators},
    phase-space samplers and matrix element generators are
    now widely available so the acquisition of data is only
    limited by time. Even then, this is a bottleneck only
    for the most expensive of processes as generation
    of data is trivial to parallelise. In the ideal case, the
    surrogate model would be faster than the matrix
    element generators, and sufficiently accurate. Neural
    network predictions are fast by virtue of them being
    simple matrix multiplications, and because they are predicted
    in batches, vectorisation is automatic\footnote{Event generators currently generate events one at a time, however, neural network predictions are still highly performant for single predictions.}.
    Therefore the difficulty
    in building a good surrogate model is in controlling
    the accuracy.

@@

    To summarise, the current state of matrix element
    emulation has to trade-off higher multiplicity processes
    with accuracy. The research in this thesis attempts to
    tackle both of these problems by exploiting the factorisation
    property of matrix elements (Sections \ref{sec:me_factorisation},
    \ref{sec:CO_factorisation}, and \ref{sec:OL_factorisation}),
    to build into the emulator the universal singular functions
    that describe the matrix elements in the soft and collinear limits
    (Sections \ref{sec:CS_dipoles} and \ref{sec:antenna_functions}).
    Since this method relies on the factorisation of matrix
    elements, we collectively refer to these family of models
    as factorisation-aware models.
    In Chapter~\ref{chapter:fame1}, we show that by using
    Catani-Seymour dipoles as a set of basis functions, tree-level
    matrix elements for $e^{+}e^{-} \rightarrow q \bar{q}$ + up to 3 gluons
    can be emulated accurately to well below 1\% per-point accuracy,
    even for the highest multiplicity. Extending this study
    to the one-loop level is done in Chapter~\ref{chapter:fame2},
    where antenna functions replace Catani-Seymour dipoles as
    the basis functions. Accuracy was observed to be of the
    order 1\% for the 5 parton case, with good scaling
    to higher multiplicities.

    The question of uncertainty of the neural network
    prediction is important as it directly contributes to the total
    theoretical uncertainty. An alternative to quantifying
    the uncertainty of the neural network prediction is to
    bypass the usage of the neural network prediction itself
    in production.
    Instead it can be used as a proxy for the exact matrix
    element in event unweighting. This was explored in
    Ref. \cite{Danziger:2021eeg},
    where it was shown that with a surrogate model of the event
    weight it was possible to use a double unweighting
    algorithm to increase the rate of event unweighting.
    In Chapter~\ref{chapter:fame3} we examine replacing
    the original emulator in said work with a factorisation-aware
    model to study the potential gain factors with a more
    accurate model. As previously mentioned, a consequence of
    this two-stage unweighting
    algorithm is that the exact event weight has to be evaluated
    at some point, meaning the uncertainty of the surrogate
    model is practically irrelevant.
\end{document}

>>>>

    Collisions at the LHC can produce hundreds of
    particles with complex final-state configurations.
    This leads to the design of the ATLAS and CMS experiments
    to contain of the order 100 million detection elements
    in an attempt to disentangle these events. With these large
    arrays, petabytes of data are recorded every year
    across all LHC experiments.
    This presents a practical challenge for data
    collection where machine learning has could provide a solution.
    Indeed, boosted decision trees have been use
    to enhance triggers \cite{CMS:2020cmk} to accept or
    reject data before being saved to disk.
    % At the LHCb
    % experiment, 70\% of all data retained have had
    % some contact with machine learning algorithms \cite{LHCb:2014set}.

@@

    In a machine learning context, the fitting of a
    model to data is termed training the model.
    Once the model is fitted, it is customary to use
    an unbiased dataset that is distinct from the training set, the test set,
    to evaluate the final model performance.

\subsection{Model of a neuron}
    In this context, a neuron is described by
    \begin{equation}\label{eqn:neuron}
        y = \phi(\mathbf{w}^{T}\mathbf{x} + b) = \phi_{\theta}(\mathbf{x})
    \end{equation}
    where the model inputs, $\mathbf{x}$, are multiplied
    by the model weights, $\mathbf{w}$, before being summed
    with a bias term, $b$. Collectively, the model weights
    and biases are the model parameters, $\theta$.
    This combination of terms is then modified
    by an activation function, $\phi$, which is carefully
    chosen to perform a non-linear transformation. The
    notation $\phi_{\theta}$ represents an activation function
    acting on $\mathbf{x}$ which has been combined with parameters $\theta$.
    A neuron is illustrated in Figure~\ref{fig:neuron}.

@@

    The output of a densely-connected neural network can
    be formulated by repeatedly applying Eq. (\ref{eqn:neuron})
    to give
    \begin{equation}\label{eqn:nn_output}
        % f(\mathbf{x}; \theta) = \boldsymbol{\phi}^{(n)}(\ldots \boldsymbol{\phi}^{(2)}(\; \boldsymbol{\phi}^{(1)}(\mathbf{x}; \theta^{(1)})\; ; \theta^{(2)}) \; ; \theta^{(n)}) \, ,
        f(\mathbf{x}; \theta) = \boldsymbol{\phi}^{(n)}_{\theta}(\; \ldots \boldsymbol{\phi}^{(2)}_{\theta}\; (\boldsymbol{\phi}^{(1)}_{\theta}(\mathbf{x}) ) \; ) \, ,
    \end{equation}
    where $\boldsymbol{\phi}^{(n)}_{\theta}$
    denotes the outputs and parameters of layer $n$,
    where the output layer is included in this notation.
    By chaining together activation functions, and especially
    non-linear activation functions, neural networks
    become good function approximators
    \begin{equation}\label{eqn:nn_approx}
        f(\mathbf{x}; \theta) \approx g(\mathbf{x}) \, .
    \end{equation}
    The universal approximation theorem states that neural networks are
    able to represent any continuous function, $g(\mathbf{x})$,
    with one hidden layer and a finite number of neurons \cite{HORNIK1991251}.
    However, it is very difficult to achieve this due to
    practical constraints such as limited data and network size.
    It is much more common to link
    together a larger number of hidden layers, leading
    to deep neural networks since better learning algorithms
    have been found for this case \cite{lu2017expressive}.

@@

    In order to quantify the performance of a network for
    a given task it is useful to define a loss function.
    For regression tasks, such as approximating a function,
    a commonly used loss function is the mean squared error
    \begin{equation}\label{eqn:MSE}
        L_{\mathrm{MSE}} = \dfrac{1}{N} \sum_{i=1}^{N} (g(\mathbf{x}_{i}) - y(\mathbf{x}_{i}; \theta))^{2} \, ,
    \end{equation}
    where $g(\mathbf{x}_{i})$ and $f(\mathbf{x}_{i}; \theta)$
    are samples from the target distribution, and model predictions
    for inputs, $\mathbf{x}_{i}$, belonging to a data set with $N$
    samples. The loss function
    encodes the discrepancy between the truth value
    and the model prediction. Therefore, the task of finding
    an acceptable set of parameters can be reframed as an optimisation
    of the loss function. It should be noted that minimisation of
    the loss function is simply a proxy for maximising neural network
    predictive accuracy. Due to the training and testing datasets being finite,
    the generalisation of the neural network is not guaranteed for the
    true underlying function from which the datasets are sampled
    from. This problem is generally referred to as overfitting.
    We will refer back to this problem in the context of
    fitting matrix elements in Section~\ref{sec:NN_ME}.

    Since neural networks are generally aimed at
    learning non-linear functions, the loss surface
    corresponding to Eq. (\ref{eqn:MSE}) is
    a function of many variables, and is likely highly
    non-convex with many local minima.
    The methods of choice for traversing these loss surfaces
    are all iterative gradient-based methods with modifications
    to improve convergence. Broadly speaking, the algorithms
    iteratively update the neural network parameters based on the
    gradient of the loss with respect to these parameters
    with some step size (or learning rate) $\eta$. The update
    rule is given by
    \begin{equation}\label{eqn:gradient_descent}
        \theta \gets \theta - \eta \cdot \nabla_{\theta} L(\mathbf{x}_{\mathrm{batch}}; \theta) \, ,
    \end{equation}
    where $\nabla_{\theta}L(\mathbf{x}_{\mathrm{batch}}; \theta)$ is the
    gradient of the loss with respect to the parameters of the
    model, averaged over the batch of inputs $\mathbf{x}_{\mathrm{batch}}$.
    The initial state of the parameters
    are usually distributed according to normal
    or uniform distributions \cite{pmlr-v9-glorot10a,he_initialiser}.
    This variation of gradient descent is referred to
    as mini-batch gradient descent because the gradient
    $\nabla_{\theta}L$ is averaged over a mini-batch of
    samples $\mathbf{x}_{\mathrm{batch}}$. Mini-batch gradient
    descent strikes a balance between stochastic gradient
    descent (estimating gradient with one sample at a time),
    and batch gradient descent (calculating gradient with
    respect to entire training set) by having an efficient
    estimation of the gradient that is accurate enough
    for practical applications. In modern machine learning
    libraries, the updating of weights in Eq. (\ref{eqn:gradient_descent})
    is carried out via matrix multiplications which are
    highly efficient on graphics-processing units (GPUs),
    and the gradients are computed numerically with the aid
    of automatic differentiation tools. For these reasons, deep
    neural networks have become much more widespread due
    to the ease of constructing and training them, as well
    as their good predictive performance for general problems.

@@

    \subsection{Optimisation of hyperparameters}
    \label{sec:hyperparameters}
    Another set of parameters that need to be optimised
    are the model hyperparameters.
    Hyperparameters are parameters that are
    not explicitly trainable parameters of the model, instead they control
    the speed and quality of the training process.
    For instance, the learning rate in Eq. ({\ref{eqn:gradient_descent}}) is
    an important hyperparameter whose initial value has
    to be chosen carefully to observe a good rate of convergence.
    Some other examples would be
    the number of hidden layers, the choice of activation
    function, or the mini-batch size.

    The choice of hyperparameters is an important
    factor during the training process as a suboptimal set
    of hyperparameters can perform significantly worse
    than a more carefully chosen set. The process of tuning
    hyperparameters is often computationally expensive for two
    reasons. Firstly, the dimensionality of parameter space to
    choose from is large, meaning the number of samples required
    to effectively explore the parameter space grows rapidly --
    this is the so-called curse of dimensionality.
    Secondly, there is no exact a priori way to evaluate the
    final performance of a neural network given a set of
    hyperparameters, meaning training has to take place for
    some steps before the final model performance can be estimated.

@@

\section{Neural networks as matrix element surrogate models}\label{sec:NN_ME}
    Motivated by the ability of neural networks to
    model non-linear functions, as well as the rise
    in GPU usage in high energy physics, it seems
    natural to leverage neural networks to build surrogate
    models for matrix elements. There are multiple advantages
    to this: the time taken to evaluate matrix elements remains
    a bottleneck in event generation, and so building a
    fast surrogate model would accelerate this process.
    Once a fast and accurate surrogate model has been
    built, it can supplement traditional matrix
    element generators to increase statistical power of the sample
    by evaluating on many more phase-space points efficiently. This
    requires the accumulated error in the surrogate model
    predictions to be lower than the statistical error in the
    integration of matrix elements, which is a key motivation
    for building an accurate surrogate model.

    The task of building a surrogate
    model for matrix elements is in principle straightforward.
    The dataset usually consists of phase-space kinematics
    in the form of four-momenta and the targets are the
    corresponding matrix elements for the process of interest,
    at the desired order in perturbation theory. As already
    discussed in Sections~\ref{sec:MC_integration} and \ref{sec:me_generators},
    phase-space samplers and matrix element generators are
    now widely available so the acquisition of data is only
    limited by available computing time. Even then, this is a bottleneck only
    for the most expensive of processes as generation
    of data is trivial to parallelise. In the ideal case, the
    surrogate model would be faster than the matrix
    element generators, and sufficiently accurate. Neural
    network predictions are fast by virtue of them being
    simple matrix multiplications, and because they are predicted
    in batches, vectorisation is automatic\footnote{Event generators currently generate events one at a time, however, neural network predictions are still highly performant for single predictions.}.
    Therefore the difficulty
    in building a good surrogate model is in controlling
    the accuracy.

@@

    To summarise, the current state of matrix element
    emulation has to trade-off higher multiplicity processes
    with accuracy. The research in this thesis attempts to
    tackle both of these problems by exploiting the factorisation
    property of matrix elements (Sections \ref{sec:me_factorisation},
    \ref{sec:CO_factorisation}, and \ref{sec:OL_factorisation}),
    to build into the emulator the universal singular functions
    that describe the matrix elements in the soft and collinear limits
    (Sections \ref{sec:CS_dipoles} and \ref{sec:antenna_functions}).
    Since this method relies on the factorisation of matrix
    elements, we collectively refer to these family of models
    as factorisation-aware models.
    In Chapter~\ref{chapter:fame1}, we show that by forming an ansatz
    out of a linear combination of Catani-Seymour dipoles with
    fitted coefficients, tree-level
    matrix elements for $e^{+}e^{-} \rightarrow q \bar{q}$ + up to 3 gluons
    can be emulated accurately to well below 1\% per-point accuracy,
    even for the highest multiplicity. Extending this study
    to the one-loop level is done in Chapter~\ref{chapter:fame2},
    where antenna functions replace Catani-Seymour dipoles as
    the ingredients in the ansatz. Accuracy was observed to be of the
    order 1\% for the 5 parton case, with good scaling
    to higher multiplicities.

    The question of uncertainty of the neural network
    prediction is important as it directly contributes to the total
    theoretical uncertainty. An alternative to quantifying
    the uncertainty of the neural network prediction is to
    use the prediction in an intermediary step of a full
    calculation, where the final prediction comes from the
    accustomed matrix element providers. This was explored in
    Ref. \cite{Danziger:2021eeg},
    where it was shown that with a surrogate model of the event
    weight it was possible to use a double unweighting
    algorithm to increase the rate of event unweighting.
    In Chapter~\ref{chapter:fame3} we examine replacing
    the original emulator in that work with a factorisation-aware
    model to study the potential gain factors with a more
    accurate model. As previously mentioned, a consequence of
    this two-stage unweighting
    algorithm is that the exact event weight has to be evaluated
    at some point, meaning the uncertainty of the surrogate
    model is practically irrelevant.
\end{document}
